Memories of migrations past: Sociality and cognition in dynamic, seasonal

### 2021-07-13 text/ms.Rmd
98% there

---
title: "Modeling memory and taxis to analyze migrations"
author: "E. Gurarie, S. Potluri, B. Fagan, others"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_float: yes
theme: cerulean
editor_options:
  chunk_output_type: inline
bibliography: ../library/memorymigration.bib
csl: ecology.csl
---

@@

@@

Much as the ultimate causes of migration vary across species, the proximate causes, drivers and mechanism can vary considerably across and even within species [@Shaw2016]. Some migrants follow a "green wave" of spring vegetation as it flowers across altitudinal or latitudinal gradients [@Bischof2012;@Kolzsch2015]. These migrations can be considered "tactical", as they can occur - as an extreme simplification - purely as  response to local conditions.   Other migrants perform long-distance migrations in anticipation that critical resources will be available at the time of arrival at the end point of migration (**refs**).  This second behavior involves the greatest trade-off between the costs of migration against the benefits of accessing resources, whether food, suitable habitat for breeding, or predator refuge, that are highly seasonal and localized (**refs**).  This approach can be considered "strategic" in the sense that it is driven not by immediate cues but by a remembered anticipation [@Bracis2017].

@@

Diffusion-advection models have a long pedigree in animal movement modeling [@Skellam1951;@Turchin1998;@Okubo2001]. These models are grounded in the general idea that animal movements - somewhat like movements of physical particles - combine a random (diffusive) component with a directed (advective) component. While direct relationships between diffusion models and movement data are somewhat tenuous [@Gurarie2011a], as a theoretical tool for exploring processes they are invaluable for their versatility and the relative ease of numeric computation of the partial differential equations (PDEs) that are used to describe diffusion-advection models mathematically. Thus, much theoretical and some applied work has been done on refining the basic assumptions of diffusion models, e.g. by including heterogeneity in populations [@Skalski2003;@Gurarie2009], fat-tailed dispersal kernels [@Kot1996], non-linear or otherwise complex responses to resources and consepecifics (**refs**).

@@

Here, we develop a diffusion-advection model with memory to explore the resilience of a migratory population under various dynamic, seasonal resource distributions.  In formulating the model, our goal is to identify the minimum set of movement and memory parameters required to generate an adaptive, migratory behavior. This includes the ability to learn to migrate from non-migratory initial conditions, simulating the release of naive animals in a seasonal environment [@Jesmer2018], or to lose the propensity to migrate if the resource distribution does not require it, which is also a commonly observed phenomenon (**refs**).  Our ultimate goal is to assess the role of long and short-term memory in the resilience (or fragility) of a migratory population against changing resource distribution dynamics, whether inter-annual variability or trends in resource distribution or phenology reflecting climate change.

We further anticipate that under many conditions a blending of *tactical* (i.e. direct response to resource availability or perception) and *strategic* (i.e. memory-driven and forward-thinking) behavior will help foragers navigate dynamic, seasonal environments.  Over-reliance on either strategy should be maladaptive.  We further anticipate that a shorter-term memory updating is needed to navigate trends in resource spatial distribution and temporal distribution (phenology), but that a longer-term reference memory is needed to navigate resource distributions that are stochastic (see also @Lin2021).

@@

$$
v_m(t, \theta_y) = \begin{cases}
0 &;  & t > t_1 &\text{and} & t \leq t_1 + dt_1 \\ 
s_{12} &;  &  t > t_1 + dt_1 &\text{and} & t \leq t_2 \\ 
0 &; & t> t_2 &\text{and} &t \leq t_2 + dt_2 \\ 
s_{21} &; &  t > t_2+dt_2 & \text{or} & t \leq t_1\\ 
\end{cases}
$$

@@

For numerical runs with climate change, we first run a simulation for 20 years to attain an equilibrium, and then begin shifting the location of the resource with a trend (slow 0.25, medium 0.5, or rapid 1 units per year), or by adding stochasticity (s.d. 3, 6, 12).  For the trend analyses, we then compute the slope of the migration location parameter against the resource drift and refer to this index as the *spatial adaptation* (SA) index.

@@

We assessed a wide range of parameter values and resource geometries and dynamics with the goal of answering the four main questions:

1. Can this model adapt to a discrete shift in peak resource location and timing?  What is the relative role of memory and sociality for adaptation?

2. Can this model acquire a migratory behavior from a non-migratory initial condition?

3. What is the role of a reference memory for dealing with stochastic resource dynamics?

4. Can this model adapt when the resource peaks shifts in space?

@@

![](figures/StabilityResults.png)
**Figure 3.** *Migration phenology matching across six model parameters. Low and high diffusion ($\epsilon = 1$ and $8$ in upper and lower panel blocks), tight, medium and loose swarms ($\lambda = 20, 50, 100$) left to right panels.  Within each of these blocks, high values of the resource following parmaeters $\alpha$ from 0 to 1000 are left to right, and higher values of the sociality parameter $\beta$ are bottom to top.  Finally, within each of the combinations of $\epsilon, \lambda, \alpha, \beta$, we show results ranging across 6 values of resource duration ($\sigma_t$, 3-15 left to right), and 6 values of resource extent ($\sigma_x$, 3-15 bottom to top), as in the zoomed in panel at bottom right. Finally, the color scheme reflects the total mismatch, i.e. the sum of the absolute differences between the migration timing and locations from the resource peak. White squares represent runs that numerically failed to estimate migration parameters.*

@@

![](figures/example2_learningtomigrate.png)
**Figure X.** *Example of model learning to migrate. The resource is a "weakly drifting" resource and the initial (year 0) condition is non migratory.  The simulation was run for 100 years, and a sampling of those years (labeled) are presented in panel a:  all years from 0 to 10, followerd by 20, 40 and 100.  Otherwise, panels are as in Figure 1.  Parameter values were $\epsilon  = 5$, $\alpha = 500$ and $\beta = 50$.*

Figure **X** illustrates the ability of the model animals to learn to migrate in a weakly drifting resource environment with a narrow pulse of resource peaking at 30 and -30 (at days 25 and 75), but a uniform distribution of resource at times 0 and 50.  In order to learn to migrate, the system needed to have a higher exploratory impulse (higher diffusion constant $\epsilon$), a stronger resource advection (higher $\alpha$) and somewhat weaker sociality (lower $\beta$).  The qualitative behavior of this process was to start drifting towards the summer resource, while slowly developing a weak pulse towards the winter resource as well. After first locking in on the summer resource, the winter migration, driven both by high diffusion and high resource following, slowly extended itself until both narrow peaks of resource can be consistently reached.

The model had, in general, a difficult time learning migration from a non-migratory initial condition.  Thus, out of 4047 successful runs, only 4 attained mismatch below 1, and 130 below 5 (Appendix Figure **Z**). Conditions that were more conducive to learning migration were pulses of *longer* duration (high $\sigma_t$), but *smaller* in scope (low $\sigma_x$), suggesting that the feedback that encourages migration needs to be compact in space but long enough in duration to lock in to the memory.

@@

**Figure 5.** *Adaptation to a steadily drifting resource. In three scenarios, the spatial coordinates of the resource drift by 0.25, 0.5, and 1 unit per year (top to bottom, respectively).  The y-axis is the index of spatial adaptation (SA), i.e. the trend of the memory-driven migration divided by the resource drift trend.  Values near 1 indicate a behavior that keeps up with climate change, values near indicate no change in migration behavior, and negative values indicate a trend that is opposite to the climate trend. We compare across spatial scales of sociality ($\lambda$ - x-axis), for low and high values resource following ($\alpha = 400$ and $100$ - orange and blue dots) and low and high values of sociality ($\beta = 100$ and 400, left and right panels). The size of the circles is proportional to the foraging efficiency of the resulting parameter combinations.  Finally, the bottom-right boxplots indicate the final year foraging efficiency against the SA index; purple and blue boxes indicate the highest values, orange and gray lower values.*

@@

As figure Z shows, higher values of resource following ($\alpha = 400$; orange circles) are nearly universally better for keeping up with climate change (SA values near 1).  Furthermore, when combined with high sociality ($\beta = 400$; right panels), nearly all parameter combinations do a good job keeping up with climate change (SA values ranging between 0.53 and 0.85 for a swarm size greater than 50). However, that maximum value is still less than 1, suggesting that truly matching a steadily drifting trend is very difficult.  Small swarms ($\lambda < 50$) have a very hard time adapting when the social attraction is high, but do fairly well when social attraction is lower.   But larger sized swarms do progressively worse across more parameterizations, e.g. in the most rapid climate change scenario, the SA drops from 0.83 to -0.13 as the swarm increases in size from 40 to 200 (essentially, the entire spatial domain).

@@

While recent memory can be helpful for adapting to a single novelty or a smoothly changing conditions, we hypothesized that more conservative approach, relying on a "reference memory" may be more beneficial when conditions change stochastically.  We tested this hypothesis by solving a set of models across a range of $\kappa$ values fom 0 (all recent memory) to 1 (all reference memory).  In these scenarios, we ran the system for as many years as needed with no stochasticity to acquire a stationary state (i.e. similarity index greater than 0.9999), used the stationary state to define the reference migration memory, and then ran the process for an additional 50 years with a stochasticity (i.e. standard deviation in peak location of the resource) ranging from 0 to 12 units.

Figure 6 illustrates that, as predicted, the highest level of $\kappa$ can significantly help foraging efficiency; however this is highly sensitive to the spatial scale of sociality.  When that scale of sociality is high enough (e..g $\lambda = 120$, blue colors) there is greater probability of overlap with a stochastic resource, and a conservative, stable migratory regime is much more beneficial in the long run.   However, if that scale is too small, then - counterintuitively - the process is overly fixed on repeatability and misses the opportunity to increase its foraging by "randomizing" its migration movements via memory of improved foraging in a previous year.  Overall, as anticipated, the greater the stochasticity, the lower the foraging efficiency; and the greater the stochasticity, the greater the benefit of a conservative reference memory driven strategy.

```{r StochasticityOnly, cache = TRUE}
require(magrittr); require(plyr); require(ggplot2); require(ggthemes); require(gtools)
a <- load("../results/stochasticity/stochasticity.rda")
b <- load("../results/trendstochasticity/trendstochasticity.rda")

cc <- smartbind(ccsigma, 
                subset(ccmusigma, psi_x == 0)) %>% 
  subset(alpha == 100 & beta == 400 & beta_x == 0)

cc.summary <- cc %>% 
  ddply(c("lambda", "kappa", "beta_x", "psi_x"), 
        summarize, 
        n = sum(!is.na(avgFE)), 
        FE_mean = mean(avgFE, na.rm = TRUE), 
        FE_sd = sd(avgFE, na.rm = TRUE)) %>% 
  mutate(FE_se = FE_sd/sqrt(n),
         FE_high = FE_mean + 2*FE_se,
         FE_low = FE_mean - 2*FE_se)

cc.summary %>% subset(lambda > 40 & psi_x > 0) %>% 
    ggplot(aes(kappa, FE_mean, col = factor(lambda), ymin = FE_low, ymax = FE_high)) +
    facet_grid(.~psi_x) + geom_point() + theme_few() + geom_errorbar() + 
    geom_path()

```

**Figure 6** *Mean and standard error of foraging efficiency across various values of reference memory $\kappa$ (x-axis) for increasing amounts of interannual stochasticity ($\psi_x$, panels left to right), and two values of spatial scale $\lambda = 80$ and $120$.  For the stochastic processes ($\psi_x > 0$), the process was run 30 times, and the mean and standard error of the foraging efficiency over a 50 year post-stabilization period is presented (points and bars). In these scenarios, the resource following parameter $\alpha = 100$ and  social attraction $\beta = 400$.*

@@

We added trends to the stochastic process described above, adding 30 years post-stabilization with slower (0.25 / year) and moderate (0.5 / year) shifts in the peak of resource, with the goal of assessing the ability of the process to keep up with climate change as a function of the reference memory parameter (Figure 7). Over-reliance on reference memory ($\kappa = 1$) by definition does not allow the system to keep up with climate change.  However, over-reliance on recent memory is also sub-optimal, and - in fact - a balancing of recent and reference memory (in our relatively narrow set of similutions with a $\kappa$ value between 0.8 and 0.9) helps the system keep up with climate change better than chasing each recent shift.   Note also, that at the highest levels of stochasticity ($\psi_x = 12$), the system fails completely to keep up with the climate change. The negative spatial adaptation index indicated that the migration behavior, in fact, can completely fall apart.

![](figures/TrendStochasticity_v2.png)

**Figure 7**. *Role of reference memory in adaptying to climate change for increasingly stochastic resource dynamics.  We ran the model with slow (mean shift 0.25 / year; blue colors) and moderate outward trend (0.5 / year; red colors), at four levels of stochasticity (inter-annual standard deviation of resource peak 0, 3 ,6 and 12, left to right panels). For non-zero stochasticity, we ran the process 30 times and present the mean and standard error of the spatial adaptation index across various values of the reference memory parameter $\kappa$: where $\kappa = 0$, the system modifies its migration based entirely on recent experiences; at $\kappa = 1$, the memory never changes from the reference memory.  Other parameter values are resource following $\alpha = 100$, social attraction $\beta = 400$, and social spatial scale $\lambda = 80$.*

@@

We found that the straightforward inclusion of a migration-specific memory in a social diffusion-advection framework was sufficient to capture many fundamental phenomena related to migration. In essence, our model allowed a population with a hybrid migratory and resource following behavior to adjust the migratory behavior based on recent experiences with the resource location. The model, while simple, was able to emulate the successful navigation of an environment with temporally and spatially isolated seasonal resource patches, the spontaneous emergence of a migratory behavior, and intrinsic robustness to changes in those environmental resources, whether steadily shifting trends or inter-annual stochasticity.  We note that our model did not include any selection, inheritance or birth or death processes, i.e. there was evolutionary optimization of parameters to maximize a fitness, as in several somewhat similar models that explicitly explore the evolution of migration [@Guttal2010;@Shaw2013;@Anderson2013].  Rather, we postulated a process defined by relatively few parameters, and explored the ability of the process to navigate a fixed set of environments with the goal of understanding the combinations of factors that lead to a resilient migration behavior. The foraging efficiency was a convenient metric of the utility of the migratory behavior, but was not a measure explicitly maximized by the model.

It is worth considering the bioloogical interpretations of the various parameters. The diffusion parameter ($\epsilon$) captures short time-scaled randomness and exploration of a population and the foraging advection strength ($\alpha$) captures the attraction of the population to better quality resources at a relatively large scale.  These are the most common ingredients in diffusion-advection models of animal movement and direct parallels can be drawn between these and, respectively, individual random movement models [@Gurarie2011] and resource selection models, which are the fundamental ingredients of empirical analysis of animal movements. The sociality parameters, which turn out to be exceedingly important for the resiliency of this process, capture two very important but lesser-studied properties of animal movements.  The strength of sociality ($\beta$) essentially quantified the importance of the group.  Because both resources and populations were normalized to 1, the ratio between $\beta$ and $\alpha$ can be interpreted as the relative importance of foraging to social cohesion; and it appears a certain balance of both is needed to successfully maintain, acquire, or adapt a migratory phenology. The spatial scale of the social group ($\lambda$) captures the spatial extent of the population, and was, somewhat unexpectedly, among the most important parameters.  In particular, a small spatial scale tended to have a much more difficult time locking in to an adaptive migratory pattern, and only in conditions where the social attraction was relatively weak. This result points to an important fundamental dynamic: that in order to adapt to any shift in resources, some individuals must experience that resource, and the two sociality-related mechanisms to have that scope is either a relatively large spatial scale or a relatively weak social cohesion.

The ability to maintain, learn and adapt migration also depends strongly on properties of the resource dynamics.  In particular, the reinforcement of memory and foraging is strongest when patches are concentrated in time, but relatively large in space.  Interestingly, in most essentially stable patterns, the eventual targeted migration arrival time coincided with the *peak*, rather than the beginning, of the resource dynamic.  Ecologically, this indicates that the long-distance social migration behavior may be particularly reinforced when the targeted resource is very sudden (explosive), which is the case for the rapid green-up that occurs in high latitudes as snow recedes in tandem with extended day lengths, leading to a rapid and intense local green-up.

Our model is somewhat similar to an evolutionary individual-based model [@Anderson2013], which showed that some inherent phenotypic plasticity is required to hedge both against persistent trends and increased stochasticity of resource distributions, but at the risk of sub-optimal access to resources [@Anderson2013].

Collective knowledge has been shown to be important, if not essential, to the evolution and process of migration [@Shaw2013;@Guttal2010;@Berdahl2018].  In our model, we assume that the migration process itself is driven by a collective trigger for the timing and locations of seasonal ranges and migration behavior.  Synchrony of migration timing and high site fidelity are well documented for migratory species [@Gurarie2019;@Joly2021] **more**.

*Some comments on systems that may or may not be more resilient to climate change based on our results*.

>>>>

---
title: 'Memories of migrations past: Sociality and cognition in dynamic, seasonal
  environments'
author: "E. Gurarie, S. Potluri, C. Cosner, R. Cantrell, W.F. Fagan"
date: "`r Sys.Date()`"
output:
  pdf_document:
    keep_tex: true
    citation_package: natbib
  html_document:
    toc: yes
    toc_float: yes
theme: cerulean
editor_options:
  chunk_output_type: inline
bibliography: ../library/memorymigration.bib
csl: ecology.csl
---

@@

# Abstract

*to do.*

@@

Much as the ultimate causes of migration vary across species, the proximate causes, drivers and mechanism can vary considerably across and even within species [@Berthold1999;@Shaw2016]. Some migrants follow a "green wave" of spring vegetation as it flowers across altitudinal or latitudinal gradients [@Bischof2012;@Kolzsch2015]. These migrations can be considered "tactical", as they can occur - as an extreme simplification - purely as response to local conditions.   Other migrants perform long-distance migrations in anticipation that critical resources will be available at the time of arrival at the end point of migration [@Abrahms2019].  This second behavior involves the greatest trade-off between the costs of migration against the benefits of accessing resources, whether food, suitable habitat for breeding, or predator refuge, that are highly seasonal and localized.  This approach can be considered "strategic" in the sense that it is driven not by immediate cues but by a remembered anticipation [@Bracis2017].

@@

Diffusion-advection models have a long pedigree in animal movement modeling [@Skellam1951;@Turchin1998;@Okubo2001]. These models are grounded in the general idea that animal movements - somewhat like movements of physical particles - combine a random (diffusive) component with a directed (advective) component. While direct relationships between diffusion models and movement data are somewhat tenuous [@Gurarie2011], as a theoretical tool for exploring processes they are invaluable for their versatility and the relative ease of numeric computation of the partial differential equations (PDEs) that are used to describe diffusion-advection models mathematically. Thus, much theoretical and some applied work has been done on refining the basic assumptions of diffusion models, e.g. by including heterogeneity in populations [@Skalski2003;@Gurarie2009], fat-tailed dispersal kernels [@Kot1996], non-linear or otherwise complex responses to resources and consepecifics.

@@

Here, we develop a diffusion-advection model with memory to explore the resilience of a migratory population under various dynamic, seasonal resource distributions.  In formulating the model, our goal is to identify the minimum set of movement and memory parameters required to generate an adaptive, migratory behavior. This includes the ability to learn to migrate from non-migratory initial conditions, simulating the release of naive animals in a seasonal environment [@Jesmer2018], or to lose the propensity to migrate if the resource distribution does not require it, also a commonly observed phenomenon [@Wilcove2008].  Our ultimate goal is to assess the role of long and short-term memory in the resilience (or fragility) of a migratory population against changing resource distribution dynamics, including both variability or trends in resource distribution or phenology reflecting climate change.

We anticipate that under many conditions a blending of *tactical* (i.e. direct response to resource availability or perception) and *strategic* (i.e. memory-driven and forward-thinking) behavior will help foragers navigate dynamic, seasonal environments.  Over-reliance on either strategy should be maladaptive.  We further anticipate that a shorter-term memory updating is needed to navigate trends in resource spatial distribution and temporal distribution (phenology), but that a longer-term reference memory is needed to navigate resource distributions that are stochastic (see also @Lin2021).

@@

$$
v_m(t, \theta_y) = \begin{cases}
0; & t > t_1  \,\, \text{and} \,\, t \leq t_1 + dt_1 \\ 
s_{12}; & t > t_1 + dt_1 \,\, \text{and} \,\,  t \leq t_2 \\ 
0; & t> t_2 \,\, \text{and} \,\, t \leq t_2 + dt_2 \\ 
s_{21}; &  t > t_2+dt_2 \,\, \text{or} \,\, t \leq t_1\\ 
\end{cases}
$$

@@

For numerical runs with climate change, we first run a simulation for 20 years to attain an equilibrium, and then begin shifting the location of the resource with a trend (slow 0.25, medium 0.5, or rapid 1 units per year), or by adding stochasticity (s.d. 3, 6, 12).  For the trend analyses, we then compute the slope of the migration location parameter against the resource drift and refer to this as the *spatial adaptation* (SA) index.

@@

We assessed a wide range of parameter values and resource geometries and dynamics with the goal of answering the four main questions: 1. Can this model adapt to a discrete shift in peak resource location and timing?  What is the relative role of memory and sociality for adaptation?  2. Can this model acquire a migratory behavior from a non-migratory initial condition?   3. What is the role of a reference memory for dealing with stochastic resource dynamics?  4. Can this model adapt when the resource peaks shifts in space?

**Table 1** *Parameters, variables and metrics.*

|| Parameter   | Description |
|-----| ----------- | ----------- |
| *Memory migration model* 
|| $\epsilon$  | Diffusion | 
|| $\alpha$  | Forage following | 
|| $\beta$  | Strength of sociality |
|| $\lambda$  | Spatial scale of sociality |
|| $\kappa$  | Proportion of reference versus working memory |
|| $x_1$, $x_2$  | location of population centroids in summer and winter| 
|| $t_1$, $dt_1$  | start and duration of summer season |
|| $t_2$, $dt_2$  | start and duration of winter season | 
|| $\kappa$ | proportion memory allocated to reference (long-term) memory vs. short-term memory|
| *Resource dynamics* 
|| $\tau$  | duration of period (year) | 
|| $m_x$, $-m_x$  | spatial coordinate of resource peak for summer and winter | 
|| $m_t$, $\tau - m_t$  | timing of resource peak for the summer and winter | 
|| $\sigma_x$, $\sigma_t$   | time duration and and spatial scale of resource pulse | 
|| $dx$, $dt$  | rate of change of peak locaton and timing of resource | 
|| $\psi_x$, $\psi_t$  | standard deviation of peak location and timing | 
| *Metrics* 
|| $MM_x$  | spatial migration mismatch |
|| $MM_t$  | temporal migration mismatch |
|| $TE$  | total mismatch |
|| $FE$  | foraging efficiency |
|| $SA$  | spatial adaptation index |

@@

![](figures/StabilityResults.png)
**Figure 3.** *Migration phenology matching across six model parameters. Low and high diffusion ($\epsilon = 1$ and $8$ in upper and lower panel blocks), tight, medium and loose swarms ($\lambda = 20, 50, 100$) left to right panels.  Within each of these blocks, high values of the resource following parmaeters $\alpha$ from 0 to 1000 are left to right, and higher values of the sociality parameter $\beta$ are bottom to top.  Within each of the combinations of $\epsilon, \lambda, \alpha, \beta$, we show results ranging across 6 values of resource duration ($\sigma_t$, 3-15 left to right), and 6 values of resource extent ($\sigma_x$, 3-15 bottom to top), as in the zoomed in panel at bottom right. The color scheme reflects the total mismatch, i.e. the sum of the absolute differences between the migration timing and locations from the resource peak. White squares represent runs that numerically failed to estimate migration parameters.*

@@

![](figures/example2_learningtomigrate.png)
**Figure 5.** *Example of model learning to migrate. The resource is a "weakly drifting" resource and the initial (year 0) condition is non migratory.  The simulation was run for 100 years, and a sampling of those years (labeled) are presented in panel a:  all years from 0 to 10, followerd by 20, 40 and 100.  Otherwise, panels are as in Figure 1.  Parameter values were $\epsilon  = 5$, $\alpha = 500$ and $\beta = 50$.*

Figure **5** illustrates the ability of the model animals to learn to migrate in a weakly drifting resource environment with a narrow pulse of resource peaking at 30 and -30 (at days 25 and 75), but a uniform distribution of resource at times 0 and 50.  In order to learn to migrate, the system needed to have a higher exploratory impulse (higher diffusion constant $\epsilon$), a stronger resource advection (higher $\alpha$) and somewhat weaker sociality (lower $\beta$).  The qualitative behavior of this process was to start drifting towards the summer resource, while slowly developing a weak pulse towards the winter resource as well. After first locking in on the summer resource, the winter migration, driven both by high diffusion and high resource following, slowly extended itself until both narrow peaks of resource can be consistently reached.

The model had, in general, a difficult time learning migration from a non-migratory initial condition.  Thus, out of 4047 successful runs, only 4 attained mismatch below 1, and 130 below 5. Conditions that were more conducive to learning migration were pulses of *longer* duration (high $\sigma_t$), but *smaller* in scope (low $\sigma_x$), suggesting that the feedback that encourages migration needs to be compact in space but long enough in duration to lock in to the memory.

@@

**Figure 6.** *Adaptation to a steadily drifting resource. In three scenarios, the spatial coordinates of the resource drift by 0.25, 0.5, and 1 unit per year (top to bottom, respectively).  The y-axis is the index of spatial adaptation (SA), i.e. the trend of the memory-driven migration divided by the resource drift trend.  Values near 1 indicate a behavior that keeps up with climate change, values near indicate no change in migration behavior, and negative values indicate a trend that is opposite to the climate trend. We compare across spatial scales of sociality ($\lambda$ - x-axis), for low and high values resource following ($\alpha = 400$ and $100$ - orange and blue dots) and low and high values of sociality ($\beta = 100$ and 400, left and right panels). The size of the circles is proportional to the foraging efficiency of the resulting parameter combinations.  The bottom-right boxplots indicate the final year foraging efficiency against the SA index; purple and blue boxes indicate the highest values, orange and gray lower values.*

@@

As figure 6 shows, higher values of resource following ($\alpha = 400$; orange circles) are nearly universally better for keeping up with climate change (SA values near 1).  Furthermore, when combined with high sociality ($\beta = 400$; right panels), nearly all parameter combinations do a good job keeping up with climate change (SA values ranging between 0.53 and 0.85 for a swarm size greater than 50). However, that maximum value is still less than 1, suggesting that truly matching a steadily drifting trend is very difficult.  Small swarms ($\lambda < 50$) have a very hard time adapting when the social attraction is high, but do fairly well when social attraction is lower.   But larger sized swarms do progressively worse across more parameterizations, e.g. in the most rapid climate change scenario, the SA drops from 0.83 to -0.13 as the swarm increases in size from 40 to 200 (essentially, the entire spatial domain).

@@

While recent memory can be helpful for adapting to a single novelty or a smoothly changing conditions, we hypothesized that more conservative approach, relying on a "reference memory" may be more beneficial when conditions change stochastically.  We tested this hypothesis by solving a set of models across a range of $\kappa$ values fom 0 (all recent memory) to 1 (all reference memory).  In these scenarios, we ran the system for as many years as needed with no stochasticity to acquire a stationary state (i.e. similarity index greater than 1-1e-6).  We then used the stationary state as the reference memory, and then ran the process for an additional 50 years with a stochasticity (i.e. standard deviation in peak location of the resource) ranging from 0 to 12.

Figure 7 illustrates that, as predicted, the highest level of $\kappa$ can significantly help foraging efficiency; however this is highly sensitive to the spatial scale of sociality.  When that scale of sociality is high enough (e..g $\lambda = 120$, blue colors) there is greater probability of overlap with a stochastic resource, and a conservative, stable migratory regime is much more beneficial in the long run.   Overall, as anticipated, the greater the stochasticity, the lower the foraging efficiency; and the greater the stochasticity, the greater the benefit of a conservative reference memory driven strategy.

![](figures/stochasticity.png)

**Figure 7** *Foraging efficiency across various values of reference memory $\kappa$ (x-axis) for increasing amounts of interannual stochasticity ($\psi_x$, panels left to right), and two values of spatial scale $\lambda = 80$ and $120$.  For the stochastic processes ($\psi_x > 0$), the process was run 30 times, and the mean and standard error of the foraging efficiency over a 50 year post-stabilization period is presented (points and bars). In these scenarios, the resource following parameter $\alpha = 100$, the  social attraction $\beta = 400$ and diffusion $\epsilon = 4$.* (NB: we are running more replicated to get stronger results / narrower error bars here)

@@

We added trends to the stochastic process described above, adding 30 years of spatial outward trend in the resource.  Our goal was to assess the ability of the process to keep up with climate change as a function of the reference memory parameter (Figure 8). Over-reliance on reference memory ($\kappa = 1$) by definition does not allow the system to keep up with climate change, leading to an adaptation index of 0. However, in many cases a balancing of recent and reference memory ($\kappa$ value between 0.6 and 0.8) in many cases was slightly but significantly better than relying entirely on recent memory.  The smaller spatial scale (in the selected parameter space) does a generally better job than the larger spatial scale at lower stochasticity.  At higher level of stochasticity, however, the larger spatial scale outperforms the smaller spatial scale, which completely loses track of the climate change.

![](figures/TrendStochasticity.png)

**Figure 8**. *Role of reference memory in adapting to climate change for increasingly stochastic resource dynamics.  We ran the model with a moderiate rate of climate change (mean shift: 0.5 / year) at five increasing levels of stochasticity (inter-annual standard deviation of resource peak 0, 3 ,6 and 12, left to right panels). For non-zero stochasticity, we ran the process 30 times and present the mean and standard error of the spatial adaptation index across various values of the reference memory parameter $\kappa$: where $\kappa = 0$, the system modifies its migration based entirely on recent experiences; at $\kappa = 1$, the memory never changes from the reference memory.  Other parameter values are resource following $\alpha = 100$, social attraction $\beta = 400$, and social spatial scale $\lambda = 80$.*

@@

We found that the straightforward inclusion of a migration-specific memory in a social diffusion-advection framework was sufficient to capture many fundamental phenomena related to migration. In essence, our model allowed a population with a hybrid migratory and resource following behavior to adjust the migratory behavior based on recent experiences with the resource location. The model, while simple, was able to emulate the successful navigation of an environment with temporally and spatially isolated seasonal resource patches, the spontaneous emergence of a migratory behavior, and intrinsic robustness to changes in those environmental resources, whether steadily shifting trends or inter-annual stochasticity. We postulated a process defined by relatively few parameters, and explored the ability of the process to navigate a fixed set of environments with the goal of understanding the combinations of factors that lead to a resilient migration behavior. The foraging efficiency was a convenient metric of the utility of the migratory behavior, but was not a measure explicitly maximized by the model.  
A few rather general conclusions can be made from our model.  One is that migration in general can be acquired, as has been demonstrated for translocated ungulates [@Jesmer2018], even in environments where resources are difficult to "surf", but that this requires a very strong attraction to resources, somewhat higher levels of exploratory behavior, and many years.  On the other hand, migrations can be considered somewhat fragile, in the sense that under certain conditions, e.g. increasing stochasticity, rapidly shifting resources, or a shift in some of the system parameters, migration can collapse down into a localized non-migratory, residential behavior, at times with only a marginal cost to the foraging efficiency index.  This sensitivity may explain why partially migratory populations composed of a mixture of resident and migratory individuals are so common and, apparently, evolutionary stable [@Berthold1999;@Chapman2011].

We also found that the ability to maintain, learn and adapt migration also depends strongly on properties of the resource dynamics. In particular, the reinforcement of memory and foraging is strongest when patches are concentrated in time, but relatively large in space.  Interestingly, in most stable patterns, the eventual targeted migration arrival time coincided with the *peak*, rather than the beginning, of the resource dynamic.  This indicates that the long-distance social migration behavior may be particularly reinforced when the targeted resource is very sudden (explosive), which is the case for the rapid green-up that occurs in high latitudes as snow recedes in tandem with extended day lengths, leading to a rapid and intense local green-up.

The sociality parameters - in particular, the spatial scale $\lambda$ - were somewhat unexpectedly, the most important for determining the resiliency of the process for stabilizing a migratory behavior.  In particular, a very small spatial scale tended to have a much more difficult time locking in to an adaptive migratory pattern, and only in conditions where the social attraction was relatively weak. In order to adapt to any shift in resources, *some* individuals must experience that resource, and the two sociality-related mechanisms to have that scope is either a relatively large spatial scale or a relatively weak social cohesion. On the other hand, overly large spatial scales compromized the ability of the process to track climate change, due to a diluting of the population's ability to concentrate over available resource patches when made available.

All of the parameters in our model have well-defined biological interpretations. The diffusion ($\epsilon$) captures short time-scaled randomness of movement, capturing the exploratory and short-term dispersive behavior. The foraging advection strength ($\alpha$) captures the attraction of the population to better quality resources at a relatively large scale.  These two parameters, the basic ingredients in diffusion-advection models of animal movement, have direct parallels to empirically estimated properties of animal behavior: diffusion is closely related to families of random walk models [@Gurarie2011] while the advective taxis is related to the step- and resource selection functions that are routinely estimated from movement datas [@Potts2020]. The spatial scale of the social group ($\lambda$) captures the spatial extent of the population, i.e. a population-level range.  The sociality parameter ($\beta$) quantifies the strength an individual's desire to approach the center of the social group. This is not a parameter that is typically measured, though it would - in principle - be possible to estimate in a manner analogous to step-selection function using only other conspecifics as opposed to environmental variables as a covariate.  Because both resources and populations were normalized to 1, the ratio between $\beta$ and $\alpha$ can be interpreted as the relative importance of foraging to social cohesion.  It does appears a certain balance of both is needed to successfully maintain, acquire, or adapt a migratory phenology.  Finally, the migration timing and location parameter can be similarly straightforwardly estimated from movement data [@Cagnacci2015;@Gurarie2019].  Thus, for example, @Gurarie2019 explicitly esimated the ranging area, timing, and seasonal range locations for migratory caribou (*Rangifer tarandus*), identifying the kind of inter-annual variation that is reflected in the stochastic scenarios explored here as well as some trends in the timing of migration.  Finally, we note that our  underlying assumption of the migration "urge" is consistent with the strong endogenous programs to migrate taht is well-documented, in particular, in birds that exhibit a seasonal restlessness known as *Zugunruhe* [@Berthold1999;@Helm2006].

Collective knowledge has been shown to be important, if not essential, to the evolution and process of migration [@Shaw2013;@Guttal2010;@Berdahl2018].  In our model, we assume that the migration process itself is driven by a collective trigger for the timing and locations of seasonal ranges and migration behavior.  Synchrony of migration timing and high site fidelity are well documented for migratory species [@Gurarie2019;@Joly2021]. We note, however, that diffusion-advection models can also be interpreted as a probabilistic description of a single individual's movement.  In this case, $\lambda$ would correspond to an individual home-range and $\beta$ would be an individual's tendency to be drawn to the center of that home range, akin to an individual migratory Ornstein-Uhlenbeck process [@Gurarie2017].

The reference memory parameter $\kappa$ is, of course, impossible to observe directly. Our model does, however, allow us to explore in an heuristic way the conditions under which a strong cultural tendency to migrate with certain fixed patterns can help a population hedge against stochasticity. In general, our model allows us to consider broadly the risks.

Importantly, our model did not include any selection, inheritance or birth or death processes, i.e. there was no evolutionary maximization of a fitness, as in many models that explicitly explore the evolution of migration [e.g. @Guttal2010;@Shaw2013;@Anderson2013]. Nonetheless, it is instructive to compare the fundamental assumptions behind such evolutionary individual-based models and a social memory-based model like ours.  For example, @Anderson2013 similarly explored the resilience of a population under selective pressure under persistent trends and increased stochasticity of a drifting optimal resource window, showing that a certain amount of heritable phenotypic plasticity is necessary to adapt successfully to climate change even at the cost of efficiency.  Our model underscores the fact that some level of resilience and adapatability can be attained with a purely cognitive process that balances sociality, long, and short term collective memory. Importantly, this knowledge must be transmitted through social and cultural, rather than genetic, pathways.  The high level of sociality among migratory animals, as well as multi-annual parent offspring bonds, are an evident pathway for that kind of transmission.  As with those evolutionary models, however, it is clear that when changes are too rapid, no amount of cognition can help entirely mitigate against adverse outcomes. Furthermore, if behaviors are not sufficiently plastic (i.e. if $\kappa$ is too close to 1), then adaptation is that much more difficult.
